Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-07-07 Notes/galaxy_delensing.tex
typos

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$).
We find that in general splitting galaxy surveys into tomographic binning can improve the correlation factor with CMB lensing by 30\%. 
Considering available survey, a galaxy survey like DES can remove about $14\%$ of the lensing signal. current generation CMB experiments will benefit from 
With the improvement of the noise CMB internal reconstruction will become more important and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (76\%) level.  
Nevertheless optical galaxy surveys, in particular if split in different tomographic bin will still play an important role even for CMB S4.

>>>>

The cosmic microwave background (CMB) polarization is a promising experimental dataset to test the inflationary paradigm and to probe the physics of the early universe. 
A particular component, the so-called B-modes, is indeed a direct signature of a prediction of inflation: the presence of gravitational waves in the early universe. However, reducing the instrumental noise in future experiments will not be enough to detect this signal. 
Secondary effects in the low redshift universe will also produce non-primordial B-modes adding confusion to the inflationary signal. In particular, the gravitational interactions of CMB photons with large scale structures will distort the primordial E-modes adding a lensing B-mode component to the primordial signal.
Removing the lensing part ('delensing') from the measurement of CMB B-modes will then be necessary to constrain the amplitude of the primordial gravitational waves.
Here we discuss the role of current and future large scale structure surveys in a multi-tracers approach to delensing that will improve the reconstruction of the lensing potential that lenses the CMB photons and, as a consequence, the delensing efficiency. 
We quantify this by the improvement due to delensing on the constraints on the inflationary tensor perturbations amplitude and shape ($r$ and $n_{t}$).
We find that, in general, splitting galaxy surveys into tomographic bins can improve the correlation factor with CMB lensing by 30\%. 
Considering available survey, a galaxy survey like DES can remove about $14\%$ of the lensing signal. current generation CMB experiments will benefit from ...
With the improvement of the noise CMB internal reconstruction will become more important and the fraction of removed lensing B-modes with CMB alone will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (76\%) level.  
Nevertheless optical galaxy surveys, in particular if split in different tomographic bin will still play an important role even for CMB S4.
